---
title: "Encodings"
output: learnr::tutorial
description: "Short tutorial about file encodings"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(mclm)
knitr::opts_chunk$set(echo = FALSE)
```

## Opening files with different encodings

In this tutorial we'll look a bit into the issue of encodings and the tools that {mclm} offers to tackle it.

```{r, eval = FALSE, echo = TRUE}
library(tidyverse) # activate tidyverse packages
library(mclm)      # activate mclm package
```

As an example, we will look at a particular file ("lssem006.txt") of the BASE corpus and build a concordance will all instances of `\b won` (words starting with *won*).

```{r open1, exercise = TRUE}
conc("data/lssem006.txt",
     r"--[(?xi)   \b won    ]--") %>% 
  print_kwic()
```

All three occurrences correspond to *won't*, but in the second and the third the apostrophe is missing or replaced by *㤼㸲* (depending on your R version). Even if it is invisible, you can see that the third occurrence corresponds to the third word to the right of the second occurrence... and there is nothing between *won* and *t*: no space, nothing visible.

So what is going on here? Undoubtedly, this was not the intention of the people creating BASE, but it turns out that *won't* in the corpus in a few cases is written as "won’t", with a curly closing single quote instead of a straight apostrophe symbol. The difference is hard to see, but `'` and `’` are two different characters. Moreover, the `’` clearly is not interpreted correctly by RStudio. It is mistakenly interpreted as 㤼㸲 or simply made invisible.

The reason is a mismatch between the character encoding that RStudio assumes to have been used in the corpus file and the character encoding that was actually used. Character encodings are conventions for the way characters are stored in a computer (most importantly, in files).
So first, which character encoding does RStudio assumes it has to use? You can ask RStudio about that with the following command from the {readr} package:

```{r locale, exercise = TRUE, exercise.eval = TRUE}
locale()
```

With this instruction, you ask which 'local settings' RStudio uses on your machine by default. The output of `readr::locale()` gives you different types of information, e.g. on the format of dates etc. What concerns us here, is the value for the setting called "Encoding". On the computer in which this tutorial was created, this value is "UTF-8". That value means that by default, when reading or writing text files, RStudio will assume the
encoding to be used is UTF-8.

In most of the corpora we will be using, this is a good assumption. Most corpora only contain ASCII characters, the representation of which is identical in most popular western encodings (ASCII, latin1, windows-1252, UTF-8). However, some corpora also use non-ascii characters (e.g. *á*, *ñ*, *ß*, *ç* or even different writing systems), in which case it is important to get the encoding right.
In any well-documented corpus, the documentation should contain information on the encoding that was used.

In the case of BASE, the intention probably was to only use ASCII characters. However, some instances of the non-ascii character `’` have found their way to the corpus and apparently the corpus files were saved using the encoding "windows-1252" (probably because the files
were originally created with Microsoft Word and then saved as text files). RStudio, assuming the encoding is UTF-8, interprets the ASCII characters correctly, but completely misinterprets the `’`. If you open the file with the R Studio editor and search for *won*, you may find that it interprets it as `?`.

All {mclm} functions that read text files, such as `freqlist()`, `surf_cooc()`, `conc()`, and `read_txt()`, support and argument `file_encoding` that allows us to specify that another encoding is to be used when reading a file.

```{r encoding, exercise = TRUE}
conc("data/lssem006.txt",
     r"--[(?xi)   \b won    ]--",
     file_encoding = "windows-1252") %>% 
  print_kwic()
```

Now the text is processed correctly. However, that doesn't solve the inconsistency in the corpus. If, for instance we want to find all instances of *won't*, irrespective of the character that is used for the apostrophe, we'll need an ad how regular expression such as `\b won ['’] t \b`.

To be clear, that regular expression is useless if we don't specify the correct encoding. So the following code needs to be fixed in order to show all instances of *won't*.

```{r encoding2, exercise = TRUE, exercise.eval = TRUE}
conc("data/lssem006.txt",
     r"--[(?xi)   \b won ['’] t \b  ]--") %>% 
  print_kwic()
```

```{r encoding2-solution}
conc("data/lssem006.txt",
     r"--[(?xi)   \b won ['’] t \b  ]--",
     file_encoding = "windows-1252") %>% 
  print_kwic()
```

In a perfect world the use of characters in the corpora would be perfectly consistent. But, well.